\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-17 jasa/paper-jasa.tex
fixed typo, clarified 1b

Figure~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Sec.~\ref{sec:mixture}.
The blue and green curves show $p_{c'}(\hat{s}_{c,c'}(\mathbf{x}))$,  the distributions for the score for the sub-classifiers for the three pair-wise comparisons of the mixture components. The red curves in Fig.~\ref{fig:p_sij} show the approximation of the density ratio obtained from those distributions.

@@

Figures~\ref{fig:1c} and \ref{fig:1d} show the approximate $\log \left( \hat{r}(\mathbf{x}) \right)$ as a function of $\mathbf{x}$ using a
2-layer neural network and a random forest for the classifier $\hat{s}(\mathbf{x})$.
The neural network provides a smoother approximation, while the random has some artifacts due to the fact that the
decision function is piece-wise constant.
%ratio when using respectively a linear model, a 2-layer neural network and a random
%forest.
The blue curves show the exact log-ratio, the green curves show
$\log \left( (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x}) \right)$
without calibration, while the red curve is the improved approximation $\log \hat{r}(\mathbf{x})$ calibrated
using histograms. Finally, the cyan curve shows the approximated log-likelihood ratio when decomposing the mixture, as
seen in Fig.~\ref{fig:p_sij}.  By leveraging the fact that densities are mixtures, the capacity
of the underlying classifiers can be more effectively focused on easier classification tasks,
resulting as expected in even more accurate approximations.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm, width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\gamma)$  for $\gamma=0.05$ and $\gamma=0$}
        \label{fig:1a}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1b.pdf}
        \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
        \label{fig:p_sij}
    \end{subfigure}

@@

Finally, we evaluate the diagnostics described in Sec.~\ref{S:diagnostics} for
this example. To aid in visualization, we restrict to a 1-dimensional slice of
the likelihood along $\alpha$ with $\beta=1$.   We consider three situations: i)
a poorly trained, but well calibrated classifier; ii) a well trained, but poorly
calibrated classifier; and iii) a well trained, and well calibrated classifier.
For each case, we employ two diagnostic tests. The first checks for independence
of $-2\log\Lambda(\theta)$ with respect to changes in the reference value
$\theta_1$ as shown in Eqn.~\ref{eq:theta_1_independence}. The second uses a
classifier to distinguish between samples from  $p(\mathbf{x}|\theta_0)$ and
samples from $p(\mathbf{x}|\theta_1)$ weighted according to $r(\mathbf{x};
\theta_0, \theta_1)$.  As discussed for Fig.~\ref{fig:5b}, statistical
fluctuations in the calibration lead to some noise in the raw approximate
likelihood. Thus, we show the posterior mean of a Gaussian processes resulting
from Bayesian optimization of  the raw approximate likelihood as in
Fig.~\ref{fig:6c}. In addition, the standard deviation of the Gaussian process
is shown for one of the $\theta_1$ reference points to indicate the size of
these statistical fluctuations. It is clear that in the well calibrated cases
that these fluctuations are small, while in the poorly calibrated case these
fluctuations are large. Moreover, in Fig.~\ref{fig:6a} we see that in the poorly
trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0,
\theta_1)$ has a significant dependence on the $\theta_1$ reference point. In
contrast, in Fig.~\ref{fig:6c} the likelihood curves vary significantly, but
this is comparable to the fluctuations expected from the calibration procedure.
Finally, Fig.~\ref{fig:6e} shows that in the well trained, well calibrated case
that the likelihood curves are all consistent with the exact likelihood within
the estimated uncertainty band of the Gaussian process.  The ROC curves tell a
similarly revealing story. As expected, the classifier is not able to
distinguish between the distributions when $p(\mathbf{x}|\theta_1)$ is weighted
by the exact likelihood ratio. We can also rule out that this is a deficiency in
the classifier because the two distributions are well separated when no weights
are applied to $p(\mathbf{x}|\theta_1)$. In both Fig.~\ref{fig:6b} and
Fig.~\ref{fig:6d} the ROC curve correctly diagnoses deficiencies in the
approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0,
\theta_1))$. Finally, Fig.~\ref{fig:6f} shows that the ROC curve in the well
trained, well calibrated case is almost identical with the exact likelihood
ratio, confirming the quality of the approximation.

>>>>

Figure~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Sec.~\ref{sec:mixture}.
The blue and green curves show $p_{c'}(\hat{s}_{c,c'}(\mathbf{x}))$,  the distributions for the score for the sub-classifiers for the three pair-wise comparisons of the mixture components. The red curves in Fig.~\ref{fig:p_sij} show the approximation of the density ratio (rescaled as $(1 +
\hat{r}(\mathbf{x}))^{-1}$) obtained from those distributions.

@@

Figures~\ref{fig:1c} and \ref{fig:1d} show the approximate $\log  \hat{r}(\mathbf{x}) $ as a function of $\mathbf{x}$ using a
2-layer neural network and a random forest for the classifier $\hat{s}(\mathbf{x})$.
The neural network provides a smoother approximation, while the random has some artifacts due to the fact that the
decision function is piece-wise constant.
%ratio when using respectively a linear model, a 2-layer neural network and a random
%forest.
The blue curves show the exact $\log {r}(\mathbf{x})$, the green curves show
$\log \left( (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x}) \right)$
without calibration, while the red curve is the improved approximation $\log \hat{r}(\mathbf{x})$ calibrated
using histograms. Finally, the cyan curve shows the approximated log-likelihood ratio when decomposing the mixture, as
seen in Fig.~\ref{fig:p_sij}.  By leveraging the fact that densities are mixtures, the capacity
of the underlying classifiers can be more effectively focused on easier classification tasks,
resulting as expected in even more accurate approximations.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm, width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\gamma)$  for $\gamma=0.05$ and $\gamma=0$}
        \label{fig:1a}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1b.pdf}
        \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))^{-1}$}
        \label{fig:p_sij}
    \end{subfigure}

@@

Finally, we evaluate the diagnostics described in Sec.~\ref{S:diagnostics} for
this example. To aid in visualization, we restrict to a 1-dimensional slice of
the likelihood along $\alpha$ with $\beta=-1$.   We consider three situations: i)
a poorly trained, but well calibrated classifier; ii) a well trained, but poorly
calibrated classifier; and iii) a well trained, and well calibrated classifier.
For each case, we employ two diagnostic tests. The first checks for independence
of $-2\log\Lambda(\theta)$ with respect to changes in the reference value
$\theta_1$ as shown in Eqn.~\ref{eq:theta_1_independence}. The second uses a
classifier to distinguish between samples from  $p(\mathbf{x}|\theta_0)$ and
samples from $p(\mathbf{x}|\theta_1)$ weighted according to $r(\mathbf{x};
\theta_0, \theta_1)$.  As discussed for Fig.~\ref{fig:5b}, statistical
fluctuations in the calibration lead to some noise in the raw approximate
likelihood. Thus, we show the posterior mean of a Gaussian processes resulting
from Bayesian optimization of  the raw approximate likelihood as in
Fig.~\ref{fig:6c}. In addition, the standard deviation of the Gaussian process
is shown for one of the $\theta_1$ reference points to indicate the size of
these statistical fluctuations. It is clear that in the well calibrated cases
that these fluctuations are small, while in the poorly calibrated case these
fluctuations are large. Moreover, in Fig.~\ref{fig:6a} we see that in the poorly
trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0,
\theta_1)$ has a significant dependence on the $\theta_1$ reference point. In
contrast, in Fig.~\ref{fig:6c} the likelihood curves vary significantly, but
this is comparable to the fluctuations expected from the calibration procedure.
Finally, Fig.~\ref{fig:6e} shows that in the well trained, well calibrated case
that the likelihood curves are all consistent with the exact likelihood within
the estimated uncertainty band of the Gaussian process.  The ROC curves tell a
similarly revealing story. As expected, the classifier is not able to
distinguish between the distributions when $p(\mathbf{x}|\theta_1)$ is weighted
by the exact likelihood ratio. We can also rule out that this is a deficiency in
the classifier because the two distributions are well separated when no weights
are applied to $p(\mathbf{x}|\theta_1)$. In both Fig.~\ref{fig:6b} and
Fig.~\ref{fig:6d} the ROC curve correctly diagnoses deficiencies in the
approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0,
\theta_1))$. Finally, Fig.~\ref{fig:6f} shows that the ROC curve in the well
trained, well calibrated case is almost identical with the exact likelihood
ratio, confirming the quality of the approximation.
